---
title: "RNAseq BRAF"
author: "Alicia Pliego"
date: "3/11/2022"
output: html_document
runtime: shiny
---

# RNAseq BRAF analysis

## Research questions

	1- Which genes are differently regulated genes in treated vs untreated in BRAF organoids?
	2- Are genes on the WNT signaling pathway upregulated in CTNNB1 mutated organoids?
	3- Are basal or classical state signatures visible explaining a drug resistance/sensitivity?

```{r setup, include=FALSE, echo=FALSE, message=FALSE}
library("DESeq2")
library(tximport)
library(plotly)

```


```{r,  echo=FALSE, warning=FALSE, message=FALSE}
#Import quantification files generated by DRAGEN from SALMON.
dirs <- list.files("/Users/aliciapliego/Projects/FMI_RNAseq/BRAF/data/")
```

```{r,echo=FALSE, warning=FALSE, message=FALSE}
quant_files <- list.files("/Users/aliciapliego/Projects/FMI_RNAseq/BRAF/data/",pattern="quant.genes.sf",recursive = TRUE,full.names = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# txi import quant files by genes from salmon into DESEQ2
txi <- tximport(quant_files,type="salmon", txOut=TRUE)
```


```{r,  echo=FALSE, warning=FALSE, message=FALSE}
metadata = read.table('/Users/aliciapliego/Projects/FMI_RNAseq/BRAF/data/BRAF_samples.txt', header = TRUE)

fig <- plot_ly(
  type = 'table',
  header = list(
    values = c("<b>Samples</b>", names(metadata)),
  align = c('left', rep('center', ncol(metadata))),
  line = list(width = 1, color = 'black'),
  fill = list(color = 'rgb(235, 100, 230)'),
  font = list(family = "Arial", size = 14, color = "white")
  ),
  cells = list(
    values = rbind(
      rownames(metadata), 
      t(as.matrix(unname(metadata)))
    ),
    align = c('left', rep('center', ncol(metadata))),
    line = list(color = "black", width = 1),
    fill = list(color = c('rgb(235, 193, 238)', 'rgba(228, 222, 249, 0.65)')),
    font = list(family = "Arial", size = 12, color = c("black"))
  ))

fig
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
dds <- DESeqDataSetFromTximport(txi, 
                                colData = metadata,
                                design <- ~Treatment)
```
### Visualizing library sizes 
Sample 32 and sample 26 have low read number. 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
counts <- colSums(assay(dds))
counts <- as.data.frame(counts)
rownames(counts) <- metadata$Sample
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}

hline <- function(y = 30000000, color = "red") {
  list(
    type = "line",
    x0 = 0,
    x1 = 1,
    xref = "paper",
    y0 = y,
    y1 = y,
    line = list(color = color)
  )
}
fig <- plot_ly( counts, x= rownames(counts), y= counts$counts,
  name = "Barplot of library sizes",
  type = "bar" )%>%
  layout(shapes = list(hline(30000000), list(type = "rect",line = list(color = "black"),
                                        x0 = 0.9, x1 = 2)), plot_bgcolor = "#e5ecf6", title = "Barplot of library sizes",
         xaxis = list(title = 'Samples'), 
         yaxis = list(title = 'Million of reads'))

fig
```

### Filtering non-expressed genes
- Remove genes with a count lower than 5
Plot the number of samples a gene is expressed in:
```{r,echo=FALSE, warning=FALSE, message=FALSE}
is_expressed <- assay(dds) >= 5
hist(rowSums(is_expressed),main="Number of samples a gene is expressed in",xlab="Sample Count")
```

It seems that genes are either expressed in all samples, or not expressed at all. Only keep genes that are expressed in at least 2 samples.

- Filter genes that are expressed in at least 2 samples.

### Visualize count distributions
Applying the boxplot function to the raw counts from our dataset reveals something about the nature of the data; the distributions are dominated by a few genes with very large counts.
```{r,echo=FALSE, warning=FALSE, message=FALSE}
keep <- rowSums(assay(dds) >= 5) >= 2
dds <- dds[keep,]
boxplot(assay(dds))

```

```{r,  echo=FALSE, warning=FALSE, message=FALSE}
boxplot(log10(assay(dds)))

```

Log2 scale data with the vsd function from DEseq2 to remove the dependence of the variance on the mean, particurarly the high variance of the logarithm of count data when the mean is low. 
```{r,include=FALSE, echo=FALSE, warning=FALSE, message=FALSE}
# Get log2 counts
vsd <- vst(dds,blind=TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Check distributions of samples using boxplots
boxplot(assay(vsd), xlab="", ylab="Log2 counts per million",las=2,main="Normalised Distributions")
# Let's add a blue horizontal line that corresponds to the median logCPM
abline(h=median(assay(vsd)), col="blue")
```

### Heatmap of the sample-to-sample distances 
I applied the dist function to the transpose of the transformed count matrix to get sample-to-sample distances.

```{r,  echo=FALSE, warning=FALSE, message=FALSE}
sampleDists <- dist(t(assay(vsd)))
```

```{r,  echo=FALSE, warning=FALSE, message=FALSE}
library(RColorBrewer)
library(pheatmap)
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(colData(dds)$Sample, colData(dds)$Status, sep="-")
colnames(sampleDistMatrix) <- colData(dds)$Name
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         col=colors)
```

```{r,  echo=FALSE, warning=FALSE, message=FALSE}
plotPCA(vsd,intgroup="Treatment")

```

```{r,  echo=FALSE, warning=FALSE, message=FALSE}
plotPCA(vsd,intgroup="Sample")
```

We show a high correlation between the group treated and untreated, except the sample 31 which is different from the two other groups.
